---
title: "Lab R2 Questions"
author: "Josh Nowak"
date: "April 9, 2018"
output: word_document
---

Please create a html document using rmarkdown and submit for grading.

```{r, echo = F, message = F, warning = F}
require(adehabitatHR)
require(sp)
require(raster)
require(purrr)
require(tidyr)
require(dplyr)
```

Let us start by creating some data.

```{r}
x <- tibble::tibble(
  ID = rep(letters[1:5], each = 5),
  Sex = rep(sample(c("M", "F"), 5, replace = T), each = 5),
  Status = 1,
  Year = rep(2011:2015, 5)
)
```

We should discuss what was done above and why it worked.

#### Excercises Part 1 (5 pts each)

* Do an operation that returns all the data from before 2013
```{r}
x %>%
  filter(Year < 2013)
```

* Now return all the data in 2014
```{r}
x %>%
  filter(
    Year == 2014,
    Sex == "M"
  )
```
* Now return the data for Males in 2014
* Return the data for Males in 2014, but report only the number of observations
of each animal
* How many animals of each sex were surveyed in each year
Option #1
```{r}
x %>%
  group_by(Year) %>%
  summarise(
    N_ind = n_distinct(ID)  
  )
```

Option #2
```{r}
count(x, Year)
```

* What are the unique values of the ID column and how many observations per (Hint: ?dplyr::n())
* Sort the result from the previous step in descending order (Hint: ?dplyr::arrange())
* Change all the values of Status to 2 where Sex is Male using group_by and mutate

## Part 2 (5 points each)

* Why did that last chunk of code work?  Why didn't we have to call group_by?  How would the result have changed if we did call group_by first?
* Simulate data for 10 animals and add a randomly assigned sex to each individual.  Make the males move more than females when simulating data.
* Summarize the movements by calculating the mean of each sex to prove that your simulation code had the desired effect.
