---
title: "Rasters"
author: "Josh"
date: "April 3, 2018"
output: 
  github_document:
    toc: true
    toc_depth: 3
---

## Review

#### Vector Data

- SpatialPoints
- SpatialLines
- SpatialPolygons
- SpatialPixels

**Slots**

- coords
- bbox
- proj4string

**Add Attributes**

- *DataFrame

**Projections**

- [spatialreference](www.spatialreference.org)
- proj4string
- EPSG

**Conversion**

- as.data.frame

**Subsetting**

- [
- subset

**Reproject**

- spTransform

**Read/Write**

- load/save
- readOGR/writeOGR
- raster/writeRaster
- GDAL via gdalUtils


**Main Packages**

- sp
- raster
- rgdal
- rgeos
- tmaps
- maps
- maptools
- rasteVis
- ggmap
- googlemaps

***

## Rasters

```{r, message=F, warning=F}
  require(raster)
```

Make our first raster

```{r}
  r <- raster()
  r
  
  ncell(r)
  r[] <- rnorm(ncell(r))
  plot(r)
  
  hasValues(r)
  inMemory(r)
```

```{r}
ncell(r)
nrow(r)
ncol(r)
xmax(r)
ymin(r)
res(r)
extent(r)

```


###  Rasters can be multi-layered
```{r}
r <- raster()
r[] <- rnorm(ncell(r))
r2 <- r * 2
r3 <- r + 100

par(mfrow = c(1, 3))
plot(r)
plot(r2)
plot(r3)

s <- stack(r, r2, r3)
```


*If your raster stack fails to build because the dimensions of the rasters are slightly different, try resample to a reference raster.*

```{r}
filename <- system.file("external/test.grd", package = "raster")
r <- raster(filename)
plot(r)
```

```{r}
r <- raster(nrow = 5, ncol = 5)
r[] <- 1:ncell(r)

msk <- raster(nrow = 5, ncol = 5)
msk[1,] <- 1
msk[,2] <- 1

new_r <- mask(r, msk)

```

Because rasters can be memory hogs it is safer to use the helper functions provided to supply values and summarise.

```{r}
r <- raster(nrow = 10, ncol = 10)
values(r) <- 1:100

#  Not super memory safe or efficient
r_not_3 <- r[r != 3]

# Use cell numbers to extract values
r[4]

get_Vals <- c(6, 3, 4, 5)

beginCluster(4)
extract(r, matrix(c(1, 1), ncol = 2, nrow = 1), cellnumbers = T)
endCluster()

```


```{r}

r3 + r2

cellStats(r3, "mean")

```

```{r}
aggregate(r2, fact = 2, fun = sum)


```

Make objects smaller to be more efficient
crop(r2, idaho_shp)

Reclassify or bin your data 
reclassify

Convert to or from sp
rasterize

Area summaries
focal
zonal

Distance
distance
gDistance
adjency

boundaries
area
clump


